The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-10 rational.tex
conclusion
_
intro

\paragraph*{Negative Types.} 
Consider the following algebraic manipulation relating a natural number {{a}}
to itself (ignoring the dotted line for a moment):
\begin{center}
\scalebox{1.1}{
\includegraphics{diagrams/thesis/algebra-wire1.pdf}
}
\end{center}
Although seemingly pointless, this algebraic proof corresponds, in our model,
to an isomorphism of type {{a <-> a}} with a non-trivial and interesting
computational interpretation. The witness for this isomorphism is a
computation that takes a value of type~{{a}} (say \$20.00) and eventually
produces the same \$20.00 value as its output. As the semantics of
Sec.~\ref{sec:rat} formalizes, this computation `flows along the dotted
line' with the following intermediate steps:
\begin{itemize}
\item We start at line (0) with \$20.00; 
\item We proceed to line (1) with the same \$20.00 but tagged as being
  in the left summand of the sum type `{{a+0}}'; we indicate this value
  as {{left 20}}
\item We continue to line (2) with the same value {{left 20}};
\item At line (3), as a result of re-association the tag on the \$20.00
  changes to indicate that it is in the left-left summand, i.e., the value is
  now {{left (left 20)}};
\item At line (4), we find ourselves needing to produce a value of type 0
  which is impossible; this signals the beginning of a reverse execution
  which sends us back to line (3) with a value {{left (right 20)}};
\item Execution continues in reverse to line (2) with the value 
  {{right (left 20)}};
\item At line (1) we find ourselves again facing an empty type so we reverse
  execution again; we go to line (2) with a value {{right (right 20)}};
\item We proceed to line (3) and (4) with the value {{right 20}};
\item We finally reach line 5 with the value {{20}}.
\end{itemize}
The example illustrates that the empty type and negative types have a
fascinating computational interpretation as values that backtrack to
satisfy dependancies and have some flavor of continuations.

\paragraph*{Fractional Types.} What is even more fascinating is that
the above computation can be realized with a completely different
concurrent schedule as follows, with fractional types.

@@

%% The execution begins at line~(0) with the value {{20}}, at line (2)
%% with the value {{right (right 20)}} flowing forwards and the value
%% {{right (left 20)}} flowing backwards, and at line (3) with the value
%% {{left (left 20)}} flowing forwards and the value {{left (right 20)}}
%% flowing backwards.

%\emph{Simultaneously} 
The execution begins at line~(0) with the value {{20}}. At line~(1)
values {{20}} and {{()}} flow forward, corresponding to the type
`{{a*1}}'. At line~(2) three independent subcomputations \emph{proceed
  concurrently}: the intial value {{20}} flowing from line (1) and two
entangled values, let us call them {{1/x}} and {{x}}, such that
specifying one implicitly specifies the dependant other. These three
subcomputations \emph{exist at the same time}. At line (3), the
dependant {{1/x}} is fixed to be the value {{20}}, thereby making
fixing values of all three branches.
%% the positive value created at line (2) which can flow to the
%% consumer; the negative value created at line (3) which can pay for the
%% debt created at line (2); and the initial value from line (0) which
%% can flow to the required forward-flowing value needed at line (3). 
At some intuitive level, one might understand the situation as
modeling a credit card payment in which money is created at line (2)
for the consumer together with a corresponding debt. The producer's
bank pays for the debt at line (3) and waits for the actual money to
arrive from the producer.  There are at least two fundamental points
about this example that must be emphasized:
\begin{itemize}
\item As the example illustrates, negative types corresponds to
  ``debts'' satified by backtracking and fractional types correspond
  to debts satisfied by constraint propagation.  It would clearly be
  disastrous if debts could be deleted or duplicated. This simple
  observation explains why these types are much simpler and much more
  appealing in a framework where information is guaranteed to be
  preserved. In previous work that used negative types (see
  Sec.~\ref{sec:related}), complicated mechanisms are typically needed
  to constrain the propagation and use of negative values because the
  surrounding computational framework is, generally speaking, careless
  in its treatment of information.
\item The appeal of negative and fractional types can already be
  gleaned from the credit card analogy. The main reason credit card
  transactions are convenient is because they disentangle the
  propagation of the resources (money) from the propagation of the
  services. Not every transaction needs both the resources and
  services to be brought together: it is sufficient to have a promise
  that the demand for resources will be somehow satisfied, as long as
  the infrastructure can be trusted with such
  promises. 
%% Computationally, this means that negative types allow us
%%   to break the dependency between producers and consumers and let each
%%   compute independently.

@@

\paragraph*{Functions and Dependancy Splitting.} 
%% Consider the following realization of boolean negation written using a
%% pattern-matching style:
%% \begin{verbatim}
%% notB b = case b of 
%%   False -> True
%%   True -> False
%% \end{verbatim}
%% In a functional language, we are not used to treating the pattern-matching
%% clauses \verb|False -> True| and \verb|True -> False| as first-class entities
%% but fractional types allow us to do just that! Informally, we can treat each
%% of the pattern-matching clauses as a first-class value of type 
%% {{(1/bool) * bool}}. 
%% A small but complete example will illustrate the main aspects of this idea.

@@

In the algebra of rational numbers we have that for each {{a}} that is
not equal to {{0}}, {{1 = (1/a) * a}}. In our framework this algebraic
identity corresponds to a type isomorphism {{1 <-> (1/a) * a}}, whose
witness maybe thought of as the `identity function' at {{a}}.  For
example, for the datatype \verb#color = R|G|B#, there are three
possible witnesses for this isomorphism mapping {{()}} to either
{{(1/R,R)}} or {{(1/Gr,Gr)}} or {{(1/B,B)}} with the particular
witness determined by further constraints encountered during the
computation.  The idea of variables that capture constraints that are
satisfied later is at the heart of logic programming and hence the
witness for the isomorphisms above is generally {{(1/alpha,alpha)}}
where~{{alpha}} is a fresh logic variable.

@@

At this point we have constructed a strange mixing of two functions;
`inputs' of one form `outputs' of the other. If {{(1/alpha1,alpha2)}}
is held by one subcomputation and {{(1/alpha2,alpha1)}} is held by
another subcomputation, these `remixed functions' form a communication
channel between the subcomputations -- unifying {{1/alpha1}} with
\verb|color| {{R}} in one subcomputation, fixes it in the
other. Depending on the choice of the type {{a}}, an arbitrary number
of bits may be communicated. The dual reading of the above, with
additive functions of the form {{0<-> (-a)+b}}, correspond to
control flow and corotuining between two subcomputations -- only one
of which `exists' at any point.

@@

@@

%% Jacques Carette, Michael Adams

>>>>

\paragraph*{Negative Types.} 
Consider the following algebraic manipulation relating a natural number {{a}}
to itself (ignoring the dotted line for a moment):
\begin{center}
\scalebox{1.1}{
\includegraphics{diagrams/thesis/algebra-wire1.pdf}
}
\end{center}
Although seemingly pointless, this algebraic proof corresponds, in our model,
to an isomorphism of type {{a <-> a}} with a non-trivial and interesting
computational interpretation. The witness for this isomorphism is a
computation that takes a value of type~{{a}} (say \$20.00) and eventually
produces the same \$20.00 value as its output. As the semantics of
Sec.~\ref{sec:rat} formalizes, this computation `flows along the dotted
line' with the following intermediate steps:
\begin{itemize}
\item We start at line (0) with \$20.00; 
\item We proceed to line (1) with the same \$20.00 but tagged as being
  in the left summand of the sum type `{{a+0}}'; we indicate this value
  as {{left 20}}
\item We continue to line (2) with the same value {{left 20}};
\item At line (3), as a result of re-association the tag on the \$20.00
  changes to indicate that it is in the left-left summand, i.e., the value is
  now {{left (left 20)}};
\item At line (4), we find ourselves needing to produce a value of type 0
  which is impossible; this signals the beginning of a reverse execution
  which sends us back to line (3) with a value {{left (right 20)}};
\item Execution continues in reverse to line (2) with the value 
  {{right (left 20)}};
\item At line (1) we find ourselves again facing an empty type so we reverse
  execution again; we go to line (2) with a value {{right (right 20)}};
\item We proceed to line (3) and (4) with the value {{right 20}};
\item We finally reach line 5 with the value {{20}}.
\end{itemize}
The example illustrates that the empty type and negative types have a
fascinating computational interpretation that has some flavor of
continuations: negative types denote values that `backtrack to satisfy
dependencies', or in other words act as `debts' that are satisfied by
the backward flow of information.

\paragraph*{Fractional Types.} 
Consider a similar algebraic manipulation involving fractional types.

@@

In the case of negatives, the dotted line indicated the `flow of
control' whereas for fractionals it indicates the `flow of
constraints'. At the heart of logic programming is the idea of
variables that capture constraints and hence it useful to trace this
computation with the analogy to logic variables in mind:

As before, the execution begins at line~(0) with the value {{20}}. At
line~(1) two values, {{20}} and {{()}}, flow forward. One can think of
the value `{{()}}' (of type {{1}}) as `having a credit card': the
`credit card' isn`t money, nor is it debt, but is the option to
generate a `credit-debt' constraint.  At line~(2) we exercise this
option and hence have three values: the initial value {{20}} flowing
from line (1) and two entangled values, {{1/alpha}} and {{alpha}}. The
{{alpha}} and {{1/alpha}} are unspecified values -- i.e. we don't yet
know how much money we need to borrow, but we do know that what is
borrowed must be what is returned. Hence {{alpha}} denotes the
presence of an unknown quantity and dually {{1/alpha}} should be
thought of as the absence of an unknown quantity.

Unlike with negative types, wherein only one value existed at a time
and the computation backtracked, what is fascinating here is that each
of the three values \emph{exist at the same time}. In other words, the
computation with fractions is realized with a completely different
concurrent schedule in which every value independently and
concurrently proceeds through its subcomputation.

At line (3), the missing unknown {{1/alpha}} is fixed to be {{20}},
thereby fixing values of all three branches i.e. we have paid our debt
of the missing information and the third subcomputation has the known
value {{20}}. After payment of our debt, as line (4) shows, we have
use of our `credit card' again.

%% At some intuitive level, one might understand the situation as
%% modeling a credit card payment in which money is created at line (2)
%% for the consumer together with a corresponding debt. The producer's
%% bank pays for the debt at line (3) and waits for the actual money to
%% arrive from the producer.

The example illustrates that fractional types also have a fascinating
computational interpretation that have some flavor of continuations:
the fractional types denote values ({{1/alpha}}) that represent
`missing information which must be supplied' in much the same sense
that continuations denote evaluation contexts with `holes that must be
filled'.  There are at least three fundamental points about the credit
card examples that must be emphasized:

\begin{itemize}
\item 
As the example illustrates, both negative types and fractional types
corresponds to ``debts'' but in different ways: negatives are satisfied
by backtracking and fractionals are satisfied by constraint
propagation.

\item 
It would clearly be disastrous if debts could be deleted or
duplicated. This simple observation explains why these types are much
simpler and much more appealing in a framework where information is
guaranteed to be preserved. In previous work that used negative types
(see Sec.~\ref{sec:related}), complicated mechanisms are typically
needed to constrain the propagation and use of negative values because
the surrounding computational framework is, generally speaking,
careless in its treatment of information.

\item 
The main reason credit card transactions are convenient is because
they disentangle the propagation of the resources (money) from the
propagation of the services. Not every transaction needs both the
resources and services to be brought together: it is sufficient to
have a promise that the demand for resources will be somehow
satisfied, as long as the infrastructure can be trusted with such
promises. This idea that dependencies can be freely decoupled and
propagated can be a powerful programming tool and we leverage this in
the construction of our novel SAT-solver (Sec. \ref{sec:sat-solver}).

%% Computationally, this means that negative types allow us
%%   to break the dependency between producers and consumers and let each
%%   compute independently.

@@

\paragraph*{Functions and Dependency Splitting.} 
%% Consider the following realization of boolean negation written using a
%% pattern-matching style:
%% \begin{verbatim}
%% notB b = case b of 
%%   False -> True
%%   True -> False
%% \end{verbatim}
%% In a functional language, we are not used to treating the pattern-matching
%% clauses \verb|False -> True| and \verb|True -> False| as first-class entities
%% but fractional types allow us to do just that! Informally, we can treat each
%% of the pattern-matching clauses as a first-class value of type 
%% {{(1/bool) * bool}}. 
%% A small but complete example will illustrate the main aspects of this idea.

@@

In the algebra of rational numbers we have that for each {{a}} that is
not equal to {{0}}, {{1 = (1/a) * a}}. In our framework this algebraic
identity corresponds to a type isomorphism {{1 <-> (1/a) * a}}, whose
witness maybe thought of as the `identity function' at {{a}}.  For
example, for the datatype \verb#color = R|G|B#, there are three
possible witnesses for this isomorphism mapping {{()}} to either
{{(1/R,R)}} or {{(1/Gr,Gr)}} or {{(1/B,B)}} with the particular
witness determined by further constraints encountered during the
computation.

 and hence the
witness for the isomorphisms above is generally {{(1/alpha,alpha)}}
where~{{alpha}} is a fresh logic variable.

@@

At this point we have constructed a strange mixing of two functions;
`inputs' of one form `outputs' of the other. If {{(1/alpha1,alpha2)}}
is held by one subcomputation and {{(1/alpha2,alpha1)}} is held by
another subcomputation, these `remixed functions' form a communication
channel between the subcomputations -- unifying {{1/alpha1}} with
\verb|color| {{R}} in one subcomputation, fixes it in the
other. Depending on the choice of the type {{a}}, an arbitrary number
of bits may be communicated. The dual reading of the above, with
additive functions of the form {{0<-> (-a)+b}}, correspond to
control flow and coroutining between two subcomputations -- only one
of which `exists' at any point.

@@

\begin{comment}

@@

\end{comment}

%% Jacques Carette, Michael Adams, Khaled
